---
title: "The Power of OOP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{power-of-oop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(s3examples)
```

R is traditionally a functional programming language, meaning that when we code
in R, we tend to think in terms of functions. For instance, given a data set,
what function can I create to plot a certain variable? Object oriented
programming introduces the concept of "classes." Classes are used extensively in
other programming languages such as Java and are fascinating as they can
enforce a specific structure or behavior of an object.

## An Example Use Case

In some instances, in R, you might want to create a function that works for a
specific type of data format. Perhaps there you are interested in a specific 
column and want to ensure your data has this column present. For instance,
let us take a look at the following dataset.

```{r example data, echo = FALSE}
ref_file <- miplicorn::miplicorn_example("reference_AA_table.csv")
alt_file <- miplicorn::miplicorn_example("alternate_AA_table.csv")
cov_file <- miplicorn::miplicorn_example("coverage_AA_table.csv")

data <- miplicorn::read_tbl_ref_alt_cov(
  ref_file,
  alt_file,
  cov_file,
  gene == "atp6" | gene == "crt"
)

data
```

Let us say we want to create a function `mutation_prevalence` which determines
the prevalence of each mutation in the dataset. We additionally wanted to plot
this function.

```{r mutation prev}
prevalence <- mutation_prevalence(data, 5)
prevalence
```

While creating this function may require some thought, it makes intuitive sense
that the result will contain four columns as seen above. Two of these columns,
`n_total` and `n_mutant`, are used to compute the final column, `prevalence`. 
Therefore, if we wanted to visualize or data, it is not very important for us
to consider the two columns `n_total` and `n_mutant`. Rather, we care about
the `mutation_name` and the `prevalence`.

If we were to create an function, `plot_prevalence`, the easiest way to code it
would be to give the function a `data` argument:

```{r basic fn, eval = FALSE}
plot_prevalence <- function(data) {
  # code for plotting goes here
}
```

Notice that in the given format, we could feed any dataset into our function.
In other words, there are no checks to make sure we are feeding in the output
of `mutation_prevalence()`. There are a couple of strategies we could employ to
solve this issue.

First, we could check that some key columns exist.

```{r key columns, eval = FALSE}
plot_prevalence <- function(data) {
  if (!"mutation_name" %in% colnames(data)) {
    stop("Missing key column!", call. = FALSE)
  }
  
  # code for plotting goes here
}
```

Another strategy would be to use classes! If we assigned a class `mut_prev` to
the output of `mutation_prevalence()`, we could easily check if the input is of
type `mut_prev`.

```{r class check, eval = FALSE}
plot_prevalence <- function(data) {
  if (!inherits(data, "mut_prev")) {
    stop("Wrong class input!", call. = FALSE)
  }
  
  # code for plotting goes here
}
```

## But Wait, There's More!

Now while using classes to solve this issue might seem like a bit of overkill.
After all, why create an entire new class when you can just have an `if()`
statement? The real power of object-oriented programming is the ability to use
**polymorphism**^[The term polymorphism has been taken from [Advanced
R](https://adv-r.hadley.nz/oo.html#oop-systems).]. What we mean by this is the
ability to use the same function for many types of input. An example of this
behavior is the base R function `print()` which behaves differently depending on
what it is printing.

```{r print behavior}
print(c(1, 2))

print(tibble::as_tibble(c(1, 2)))
```

Using OOP, the developer can customize how certain functions behave on certain
objects. Revisiting our `mut_prev` class, we could even change the way the table
is printed! Another common example is plotting a dataset. Say we have developed
a package that introduces three different type of datasets. We can then create
a plot method for each type of class.

```{r print methods, eval = FALSE}
plot.class1 <- function(x) {
  # plotting code
}

plot.class2 <- function(x) {
  # plotting code
}

plot.class3 <- function(x) {
  # plotting code
}
```

Revisiting our class, `mut_prev`, when we call `plot()` it creates a custom plot
specific to only our class!

```{r plot class}
plot(prevalence)
```

## Additional Reading

There is a lot more to the world of OOP and many important packages leverage
various OOP techniques. In fact, in R, there are even multiple OOP systems. For
example, the [Tidyverse](https://www.tidyverse.org/) is built using the S3
system whereas the [Bioconductor project](https://www.bioconductor.org/) uses
S4. Of all the available systems, S3 is regarded as the easiest system to learn.

To learn more about OOP, I would highly recommend reading the OOP chapter of
[Advanced R](https://adv-r.hadley.nz/oo.html) and visiting some of the articles
in the [`{vctrs}` package](https://vctrs.r-lib.org/articles/s3-vector.html). The
code for this post can be found in my [`{s3examples}` package]() where I have
written a simple S3 subclass of the `tibble()` subclass.
